---
title: |
  | Olfactory camouflage and communication in birds
  |
  | R code for the comparative analysis 
  | on seasonal differences in preen oil composition
author: "*L.A. Grieves, M. Gilles, I.C. Cuthill, T. Székely, E.A. MacDougall-Shackleton, B.A. Caspers*"
output: 
  pdf_document: 
    toc: true # to generate a table of contents
    toc_depth: 3  # up to three depths of headings (specified by #, ## and ###)
    number_sections: true  ## if you want number sections at each table header
    latex_engine: xelatex
  html_document: default
header-includes: 
 \usepackage{float}
 \floatplacement{figure}{H}
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE) 
# settings for the whole documents: no message and no warning will be displayed
```

>

>

>

Required packages

```{r results='hide'}
library(ape)
library(MCMCglmm)
library(coda)
library(plotMCMC)
```

>

# Data

```{r, echo=FALSE}
# Set working directory
setwd("C:/Users/localadmin/Documents/PhD SMELLY PLOVERS/Review/Collaborative review/seasonal differences/analysis")
```

## Data on seasonal differences

Import data on seasonal differences

```{r}
df <- read.csv("data_season.csv")
# Remove rows where sex is undefined (NA) as this information is needed for the analysis
df <-df[!is.na(df$sex),]
```

Sample size = number of occurences (1 occurence = 1 sex within a species)
```{r}
nrow(df)
```

Number of occurences in each category of each variable
```{r}
ss.seas <- data.frame(table(df$seasonalchange)["0"],table(df$seasonalchange)["1"])
colnames(ss.seas) <- c("no","yes")
rownames(ss.seas) <- c("seasonal difference")
ss.seas

ss.inc <- data.frame(table(df$incubation)["0"],table(df$incubation)["1"])
colnames(ss.inc) <- c("no","yes")
rownames(ss.inc) <- c("incubation")
ss.inc

ss.nest <- data.frame(table(df$nest)["0"],table(df$nest)["1"])
colnames(ss.nest) <- c("nonground","ground")
rownames(ss.nest) <- c("nest ecology")
ss.nest

ss.timesc <- data.frame(table(df$timescale)["0"],table(df$timescale)["1"])
colnames(ss.timesc) <- c("within breeding","across breeding and nonbreeding")
rownames(ss.timesc) <- c("time scale")
ss.timesc
```

Number of species
```{r}
length(unique(df$species))
```

Number of studies
```{r}
length(unique(df$study))
```

## Phylogenetic data

Import phylogenetic data 

```{r, results='hide'}
# birdtree.org, 1000 trees, from source: Stage2 MayrAll Hackett
tree <- read.nexus("data_phylo.nex") 
# this tree contains all species included in the review
class(tree)
```

Build consensus tree

```{r, results='hide'}
# Concensus tree: given a series of trees, this function returns the consensus tree. 
# By default, the strict-consensus tree is computed. 
# To get the majority-rule consensus tree, use p = 0.5. 
# Any value between 0.5 and 1 can be used.
consensustree <- consensus(tree, p = 0.5, check.labels = TRUE)
class(consensustree)

# Check that the tree covers all the species
species <- df$phylo
setdiff(species, as.character(consensustree$tip.label)) 
# 0 species listed in our data but not in the tree
setdiff(as.character(consensustree$tip.label),species) 
# species listed in the tree but not in our data (df = data with seasonal differences)

# Trim tree to have only species where seasonal differences are known
subtree <- drop.tip(consensustree, tip=setdiff(consensustree$tip.label,df$phylo))

# Check that the tree covers all the species
setdiff(species, as.character(subtree$tip.label)) 
# 0 species listed in our data but not in the tree
setdiff(as.character(subtree$tip.label),species) 
# 0 species listed in the tree but not in our data

# Dealing with polytomies
is.binary.tree(subtree) # FALSE > there are some polytomies
subtree_random <- multi2di(subtree,random=TRUE) 
# randomization approach to take care of the polytomies
is.binary.tree(subtree_random) # TRUE > no more polytomies

```

Some name synonyms had to be fixed to match the list of bird tree: Poecile > Parus, Chlorus > Carduelis, Acanthis > Carduelis
NB: Cory's shearwater (*Calonectris borealis*) is considered a subspecies of Scopoli's shearwater (*C. diomedea*) by some authors. In birdtree.org Cory's shearwater (*Calonectris borealis*) is absent, so we used the name of Scopoli's shearwater (i.e. Calonectris diomedea) as they are phylogenetically close/equivalent

Plot tree

```{r}
# Plot tree
plot(subtree_random, cex=.7, label.offset =.1, no.margin = TRUE)
```

Get phylogenetic matrix

```{r results='hide'}
# Compute branch lengths of tree
phylo_branch <- compute.brlen(subtree_random, method = "Grafen", power = 1)

# Check tree is ultrametric
is.ultrametric(phylo_branch)

# Calculate the inverse of the matrix of phylogenetic relatedness (to be used in MCMCglmm)
inv.phylo <- inverseA(phylo_branch,nodes="TIPS",scale=TRUE)
```


# MCMCglmm

## Prior

Because we have no a priori belief on the value of the parameter, we set a weakly informative prior (inverse-gamma)

```{r}
prior2 <- list(G = list(G1 = list(V = 1, n = 0.002), 
                        G2 = list(V = 1, n = 0.002)), 
                        R = list(V = 1, n = 0.002))
```

## Model

Model parameters: burnin and thin should be big enough to get rid of autocorrelation and ensure convergence, nitt should be big enough to have sufficient sample size (nitt-burnin/thin >=1000).

```{r,cache=TRUE}
# Model 1
mcmc1 <- MCMCglmm(seasonalchange ~ incubation * nest + timescale, 
                  random = ~ phylo + species,
                  family = "categorical", 
                  data = df, 
                  prior = prior2, 
                  ginverse =  list(phylo = inv.phylo$Ainv),
                  nitt=13000000,burnin=10000,thin=3500,
                  verbose = FALSE)
```


# Model verifications

Diagnostics

```{r}
raftery.diag(mcmc1) # tells you how many MCMC iterations you need to run for convergence
```

## Check autocorrelation

Check that autocorrelation is low 

```{r}
autocorr(mcmc1$Sol[,1]) 
autocorr(mcmc1$VCV)
```

<.05 after Lag 10000 = low autocorrelation

Visualize autocorrelation

```{r}
par(mar=c(5,5,4,4))
autocorr.plot(mcmc1$Sol[,1]) 
autocorr.plot(mcmc1$VCV)
```

No pattern of autocorrelation, as it curves quickly to 0.

View traces to make sure there are no patterns or autocorrelation

```{r}
allChains <- as.mcmc(cbind(mcmc1$Sol,mcmc1$VCV))
plotTrace(allChains)
logChain<-log10(mcmc1$VCV)
plotTrace(logChain)
```

## Check convergence

Plot the trace and density of the inferred intercepts and variances 
```{r}
par(mar=c(4,4,2,2))
plot(mcmc1$Sol) # intercept
plot(mcmc1$VCV) # variance
```

Looks like a "hairy caterpillar, suggesting good convergence.

## Gelman-Rubin Statistics

Comparison of the same model run multiple times

Run the same model two more times

```{r, cache=TRUE}
# Model 2
mcmc2 <- MCMCglmm(seasonalchange ~ incubation * nest + timescale, 
                  random = ~ phylo + species,
                  family = "categorical", 
                  data = df, 
                  prior = prior2, 
                  ginverse =  list(phylo = inv.phylo$Ainv),
                  nitt=13000000,burnin=10000,thin=3500,
                  verbose = FALSE)

# Model 3
mcmc3 <- MCMCglmm(seasonalchange ~ incubation * nest + timescale, 
                  random = ~ phylo + species,
                  family = "categorical", 
                  data = df, 
                  prior = prior2, 
                  ginverse =  list(phylo = inv.phylo$Ainv),
                  nitt=13000000,burnin=10000,thin=3500,
                  verbose = FALSE)
```

Compare posterior mode and mean for the three different runs 
(mean and posterior mode of each model should be similar)

```{r}
m1 <- mean(mcmc1$Sol[,1])
pm1 <- posterior.mode(mcmc1$Sol[,1])
m2 <- mean(mcmc2$Sol[,1])
pm2 <- posterior.mode(mcmc2$Sol[,1])
m3 <- mean(mcmc3$Sol[,1])
pm3 <-posterior.mode(mcmc3$Sol[,1])
stat <- c("mean","post.mode")
mcmc.1 <- c(m1,pm1)
mcmc.2 <- c(m2,pm2)
mcmc.3 <- c(m3,pm3)
data.frame(stat,mcmc.1,mcmc.2,mcmc.3)
```

Posterior modes and means are similar across runs

Gelman-Rubin statistics: the potential scale reduction factor (PSRF) should be less than 1.1 among chains

```{r}
# To check for independent chain posterior convergence
model.all <- mcmc.list(mcmc1$Sol,mcmc2$Sol,mcmc3$Sol)
gelman.diag(model.all, confidence = 0.95, transform=FALSE, autoburnin=TRUE)
```

We can see that these are very close to 1, suggesting good convergence. 

Gelman plots

```{r}
par(mar=c(3,5,3,2))
gelman.plot(model.all)
```

This is also evident when looking at the plot of the values generation per generation as the mixing is very good.


# Results

Summary of the three models

```{r}
summary(mcmc1)
summary(mcmc2)
summary(mcmc3)
```

The effective sample sizes are all relatively good (higher than the value recommended by the function raftery.diag), both for random and fixed effects.

The three runs give the same results but values differ slightly. The second and third models were run only for verification purposes (Gelman-Rubin statistics). We report only the results of the first model (i.e. mcmc1).

## Fixed effects

Plot the parameter estimates and 95% CIs

```{r echo=FALSE}
plot.estimates <- function(x) {
  if (class(x) != "summary.mcmc")
    x <- summary(x)
  n <- dim(x$solutions)[1]
  plot(x$solutions[,1], n:1,
       yaxt="n", ylab="",
       xlim=range(max(x$solutions[,3])*1.02, min(x$solutions[,2])*1.02),
       pch=19,
       main="Posterior means and 95% credible intervals")
  grid()
  axis(2, at=n:1, rownames(x$solutions), las=2)
  arrows(x$solutions[,2], n:1, x$solutions[,3], n:1, code=0)
  abline(v=0, lty=2)
}

par(mar=c(5,6,4,4))
```

```{r}
plot.estimates(mcmc1)
```

## Random effects

See below.


# Random effects

We included phylogeny and species as random effects. Let's see if they have an effect on the probability to dtect seasonal differences in preen oil composition.

## Posterior distribution

Plot the posterior distribution as a histogram to check for significance and whether it's been well estimated or not. 

```{r}
par(mfrow = c(1,2),mar = c(6, 6, 2, 2)) # plot panel
hist(mcmc(mcmc1$VCV)[,"phylo"]) # non-significant
hist(mcmc(mcmc1$VCV)[,"species"]) # non-significant
```

```{r echo=FALSE}
par(mfrow=c(1,1)) # Reset the plot panel back to single plots
```

Posterior distribution is pushed up against 0 for both random effects (phylogenetic relatedness and species), which therefore seem to have little effect on seasonal variation in preen oil composition. Indeed, because variance cannot be zero, if the mean value is pushed up against zero, it reflects a non-significant effect. The large spread of the histogram indicate that the distribution is not well estimated (poor precision).

## Proportion of the total variance explained by the random effects

To evaluate variance explained by the random effects, it is useful to look at the relative proportion of the variance explained by each effect and by the residuals (units).

```{r}
# Proportion of variance explained by random factors
rand <- mcmc1$VCV/apply(mcmc1$VCV,1,sum)
# Get median values (50%) and 95% quantiles
apply(rand,2,function(c) quantile(c,probs = c(0.025,0.5,0.975)))
```
```{r}
# Get the mean value
apply(rand,2,mean)
```

The large confidence intervals confirm the poor precision of the estimates.

## Inclusion of random effects and fit of the model 

Run the different models (phylo only, species only, no random effect) to check if their inclusion in the model improves its fit. Note that a specific prior is built for each model.

Model phylo only

```{r, cache=TRUE}
prior1 <- list(R=list(V=1,nu=0.002),G=list(G1=list(V=1,nu=0.002)))

mcmc.phylo <- MCMCglmm(seasonalchange ~ incubation * nest + timescale, 
                  random = ~phylo,
                  family = "categorical", 
                  data = df, 
                  prior = prior1, 
                  ginverse =  list(phylo = inv.phylo$Ainv),
                  nitt=13000000,burnin=10000,thin=3500,
                  verbose = FALSE)

summary(mcmc.phylo)                  
```

Model species only

```{r, cache = TRUE}
mcmc.species <- MCMCglmm(seasonalchange ~ incubation * nest + timescale, 
                  random = ~species,
                  family = "categorical", 
                  data = df, 
                  prior = prior1, 
                  ginverse =  list(phylo = inv.phylo$Ainv),
                  nitt=13000000,burnin=10000,thin=3500,
                  verbose = FALSE)
summary(mcmc.species) 
```

Model no random effect

```{r,cache=TRUE}
prior0 <- list(R=list(V=1,nu=0.002))

mcmc.norandom <- MCMCglmm(seasonalchange ~ incubation * nest + timescale, 
                  family = "categorical", 
                  data = df, 
                  prior = prior0, 
                  ginverse =  list(phylo = inv.phylo$Ainv),
                  nitt=13000000,burnin=10000,thin=3500,
                  verbose = FALSE)

summary(mcmc.norandom) 
```

Compare the different models using the Deviance Information Criterion (DIC; lower values are best).

```{r}
# Compare fit of the models (DIC)
data.frame(models=c("mcmc.norandom","mcmc.phylo","mcmc.species","mcmc1"),
           random.effects=c("NA","phylo","species","phylo+species"),
           DIC=c(mcmc.norandom$DIC,mcmc.phylo$DIC,mcmc.species$DIC,mcmc1$DIC))
```

Inclusion of both phylogeny and species increases the fit of the model (lower DIC).


# Focus on the interaction nest:incubation

Split the data into nonground-nesting and ground-nesting species and run separate analyses for the effect of incubation.

Split the data

```{r, results="hide"}
# Data with only ground nesting species
df_g <- df[which(df$nest==1),]

# Data with only nonground nesting species
df_ng <- df[which(df$nest==0),]
```

Prepare phylogenetic matrices corresponding to both data sets

```{r, results="hide"}
# Trim consensus tree
subtree_g <- drop.tip(consensustree, tip=setdiff(consensustree$tip.label,df_g$phylo)) # ground nesting species
subtree_ng <- drop.tip(consensustree, tip=setdiff(consensustree$tip.label,df_ng$phylo)) # nonground nesting species

# Check that the trees cover all the species
# Ground nesting species
species_g <- df_g$phylo
setdiff(species_g, as.character(subtree_g$tip.label)) 
setdiff(as.character(subtree_g$tip.label),species_g) 
# Nonground nesting speices
species_ng <- df_ng$phylo
setdiff(species_ng, as.character(subtree_ng$tip.label)) 
setdiff(as.character(subtree_ng$tip.label),species_ng) 

# Dealing with polytomies
# Ground nesting species
is.binary.tree(subtree_g) # FALSE > there are some polytomies
subtree_g_random <- multi2di(subtree_g,random=TRUE) # randomization to take care of the polytomies
is.binary.tree(subtree_g_random) # TRUE > no more polytomies
# Nonground nesting speices
is.binary.tree(subtree_ng) # FALSE > there are some polytomies
subtree_ng_random <- multi2di(subtree_ng,random=TRUE) # randomization to take care of the polytomies
is.binary.tree(subtree_ng_random) # TRUE > no more polytomies

# Compute branch lengths of tree
phylo_branch_g <- compute.brlen(subtree_g_random, method = "Grafen", power = 1)
phylo_branch_ng <- compute.brlen(subtree_ng_random, method = "Grafen", power = 1)

# Calculate the inverse of the matrix of phylogenetic relatedness (to be used in MCMCglmm)
inv.phylo_g <- inverseA(phylo_branch_g,nodes="TIPS",scale=TRUE)
inv.phylo_ng <- inverseA(phylo_branch_ng,nodes="TIPS",scale=TRUE)
```

Model on ground nesting species 

```{r, cache = TRUE}
mcmc_g <- MCMCglmm(seasonalchange ~ incubation + timescale, 
                  random = ~ phylo + species,
                  family = "categorical", 
                  data = df_g, 
                  prior = prior2, 
                  ginverse =  list(phylo = inv.phylo_g$Ainv),
                  nitt=13000000,burnin=10000,thin=3500,
                  verbose = FALSE)
summary(mcmc_g)
```

Model on nonground nesting species 

```{r, cache = TRUE}
mcmc_ng <- MCMCglmm(seasonalchange ~ incubation + timescale, 
                  random = ~ phylo + species,
                  family = "categorical", 
                  data = df_ng, 
                  prior = prior2, 
                  ginverse =  list(phylo = inv.phylo_ng$Ainv),
                  nitt=13000000,burnin=10000,thin=3500,
                  verbose = FALSE)
summary(mcmc_ng)
```
